---
title: "Assignment 4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Please fork this to your repo (or copy it there). Make sure your repo is private, but shared with the course admin. Complete the assignment within your forked version of this file. Please start early, because this assignment is challenging and if you wait until the last moment, you will not be able to complete it. 


Group: 19
  
Group members: Siqi, Arnav, Ruini

Assignment due date: November 30, 2021, 11:59PM
 
## Introduction
*Epstein-Barr Virus (EBV) is a herpes virus and the cause of mononucleosis. Like other herpes viruses, infected cells can remain in a latent phase in the infected person their entire lives. B-cells are the usual target of EBV. When infected, B-cells can become immortalized, and can be cultured in vitro as Lymphoblastoid Cell Lines (LCLs). Like some other viruses, this immortalization can predispose the individual to certain types of cancer. For instance, Hodgkin's Lymphoma is derived from EBV-infected B cells. *

*Today, we're going to look at gene expression data to see how EBV infection changes the transcriptome. We will be looking at B-cell and LCL RNA-seq data from 5 donors, where LCLs were made from each of the donor's B-cells. Please start from this file and add code/code blocks/and commentary as needed. *


## Load data
*Load needed libraries (install any you don't yet have - edgeR can be installed through BioConductor)*
```{r pressure, echo=FALSE}
library(ggplot2)
library(edgeR)
library(reshape)
set.seed(416) # please do not change the seed or set a new seed.
```

*Load data:*
```{r}
rnaSeqData =  read.table(textConnection(readLines(gzcon(url(
  "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE126nnn/GSE126379/suppl/GSE126379_BLCL_processed.txt.gz")))), 
  sep="\t", stringsAsFactors = FALSE, header = TRUE)

```

*Reshape and re-label data to make it more usable.*

```{r}
#melt the data frame so that there is one row for every gene-sample-quantification.type (see below) combination
rnaSeqDataMelted = melt(rnaSeqData, id.vars=c("Chromosome", "Gene.Name", "Gene.ID", "Sum.of.Exon.Length.of.Gene"))
#label every value as a read count, but...
rnaSeqDataMelted$quantification.type = "counts"; 
#relabel those that are really RPKM
rnaSeqDataMelted$quantification.type[grepl(rnaSeqDataMelted$variable,pattern = "RPKM")] = "RPKM"; 
#get the donor (human) ID
rnaSeqDataMelted$donor = gsub(".*BLCL_([ABCDE])([12]).*", "\\1", rnaSeqDataMelted$variable)
#Get the identifier they use to distinguish LCLs from B-cells
rnaSeqDataMelted$celltype = gsub(".*BLCL_([ABCDE])([12]).*", "\\2", rnaSeqDataMelted$variable)
#They didn't label the cell types clearly, so we're going to guess their assignment and verify below
rnaSeqDataMelted$celltype[rnaSeqDataMelted$celltype=="1"]="B"
rnaSeqDataMelted$celltype[rnaSeqDataMelted$celltype=="2"]="LCL"
#Remove this, since we have already parsed out the needed info 
rnaSeqDataMelted$variable=NULL;
rnaSeqDataMelted$Gene.Name=NULL; #we only use Gene.ID
rnaSeqDataMelted$sampleID = paste(rnaSeqDataMelted$celltype, rnaSeqDataMelted$donor, sep="_")
#divide data into read counts and RPKM
rnaSeqDataMeltedCounts = rnaSeqDataMelted[rnaSeqDataMelted$quantification.type=="counts",]
rnaSeqDataMeltedRPKM = rnaSeqDataMelted[rnaSeqDataMelted$quantification.type=="RPKM",]
#relabel "value" as appropriate, and remove the 'quantification.type' column
rnaSeqDataMeltedCounts$counts=rnaSeqDataMeltedCounts$value; 
rnaSeqDataMeltedCounts$value=NULL; 
rnaSeqDataMeltedCounts$quantification.type=NULL;
rnaSeqDataMeltedRPKM$RPKM=rnaSeqDataMeltedRPKM$value; 
rnaSeqDataMeltedRPKM$value=NULL; 
rnaSeqDataMeltedRPKM$quantification.type=NULL;
rm('rnaSeqDataMelted') # remove 

```

*Here, `rnaSeqDataMeltedRPKM` contains the gene expression "Reads per kilobase of transcript per million reads", and `rnaSeqDataMeltedCounts` contains the raw count data.*

## Questions

### Question 1
*It is always important to perform Quality Control on your data. *

#### Question 1a (2 pts)
*For some reason, the authors labeled the different cell types "1" and "2" in this table. We need to figure out which is which. You know that EBV induces the NF-kappaB pathway, and that the TNF gene should be highly induced. Make a graph that shows the relative gene expression levels of TNF (y axis) for both cell types (x axis), for each donor (colour or "fill").*


*Enter your answer below: *
```{r}
TNF_exp<-rnaSeqDataMeltedRPKM[rnaSeqDataMeltedRPKM[,2]=="TNF" & rnaSeqDataMeltedRPKM[,7]>0,]
TNF_exp$RPKM<-log(TNF_exp$RPKM+1)/log(2)
ggplot(TNF_exp, aes(x=celltype, y=RPKM, fill=celltype))+geom_boxplot()
```

*Explain what you see. Are the celltype labels correct?*
### The cell type labels is wrong, because in this label, B cells have much higher TNF expression levels than LCL. Actually, the TNF gene should be highly induced in LCL. So 1 should be LCL and 2 should be B cells.

#### Question 1b ( 2pts)
*With RNA-seq data, it is important both that samples have sufficient coverage, and that the samples have similar coverage. Either case can lead to underpowered analysis, or misleading results. Calcualte the read coverage for each sample. Make a plot with read coverage on the y-axis (total number of reads) and the samples on the x-axis.*

```{r}
###Since the labels are wrong, we need set the correct ones
rnaSeqDataMeltedRPKM$celltype2<-rep("LCL",nrow(rnaSeqDataMeltedRPKM))
rnaSeqDataMeltedRPKM$celltype2[rnaSeqDataMeltedRPKM$celltype=="LCL"]<-"B"
rnaSeqDataMeltedRPKM$celltype<-NULL
colnames(rnaSeqDataMeltedRPKM)[7]<-"celltype"
rnaSeqDataMeltedRPKM$sampleID<-paste(rnaSeqDataMeltedRPKM$celltype,rnaSeqDataMeltedRPKM$donor,sep="_")
rnaSeqDataMeltedRPKM$sampleID<-paste(rnaSeqDataMeltedRPKM$celltype,rnaSeqDataMeltedRPKM$donor,sep="_")
rnaSeqDataMeltedCounts$celltype2<-rep("LCL",nrow(rnaSeqDataMeltedCounts))
rnaSeqDataMeltedCounts$celltype2[rnaSeqDataMeltedCounts$celltype=="LCL"]<-"B"
rnaSeqDataMeltedCounts$celltype<-NULL
colnames(rnaSeqDataMeltedCounts)[7]<-"celltype"
rnaSeqDataMeltedCounts$sampleID<-paste(rnaSeqDataMeltedCounts$celltype,rnaSeqDataMeltedCounts$donor,sep="_")

library(dplyr)
coverage<-rnaSeqDataMeltedCounts %>% group_by(sampleID) %>% summarize("Total number of reads"=sum(counts))
ggplot(coverage, aes(x=sampleID, y=`Total number of reads`,color="green")) +geom_bar(stat="identity", fill="white")+theme(legend.position="none")
```

*Which sample has the least coverage?*
### LCL_E sample has the least coverage.

```{r}
coverage$sampleID[which.min(coverage$`Total number of reads`)]
```

*Which sample has the most?*


```{r}
coverage$sampleID[which.max(coverage$`Total number of reads`)]
```
### B_D has the most.

*What is the % difference between the max and min (relative to the min)?*


```{r}
(max(coverage$`Total number of reads`)-min(coverage$`Total number of reads`))/min(coverage$`Total number of reads`)
```
### the % difference is 186%.

*In cases where samples have vastly different coverage, you can potentially down-sample the higher-coverage samples. Sometimes, throwing out the data in this way can also introduce new problems, so we're going to stick with the data we have. What is more important in our analysis is that the LCLs and B-cells have similar coverage.*

#### Question 1c ( 4.5 pts)
*One easy way to compare the samples, is to calculate the pairwise correlation between them. In general, we do this in log gene expression space.*

*First we need to create a RPKM matrix. You can use the `reshape` package to change the shape of a `data.frame`. Below is an example for how to reshape our RPKM `data.frame` into a matrix of genes by samples. *
```{r}
# Here, we need fun.aggregate because there are duplicate gene names. We'll just add everything up for each gene name.
rpkmMatrix = cast(rnaSeqDataMeltedRPKM, Gene.ID ~ sampleID, value="RPKM", fun.aggregate="sum")
#now convert this to a matrix
row.names(rpkmMatrix) = rpkmMatrix$Gene.ID
rpkmMatrix$Gene.ID=NULL;
rpkmMatrix= as.matrix(rpkmMatrix)

head(rpkmMatrix)

```

*Calculate the pairwise Pearson correlations between samples. Use log(RPKM+1) to calculate the correlations. Plot the correlations as a heatmap.*
```{r}
rpkmMatrix<-log(rpkmMatrix+1)
correlation<-cor(rpkmMatrix, method = "pearson")
library(pheatmap)
pheatmap(correlation)
```

*How similar are the B-cell samples to each other? *
### B-cell samples are very similar to each other.

*The LCLs to each other? *
### LCLs also are very similar to each other.

*The B-cells to the LCLs?*
### B-cells are very different to LCLs.

*The donors are labeled [A-E]. What would we expect to see if there were donor-specific effects here (e.g. some donor-specific signal affecting both cell types)? Do we see them?*
###If there were donor-specific effects, we would expect to see higher correlation between bothe cell types in the donor, or the donor would not cluster well with other donors, however, I do not see obvious evidence.

#### Question 1d (4 pts)

*Plot the kernel density estimate for RPKM (x axis). Describe the distribution. *

```{r}
kernel_d<-density(rnaSeqDataMeltedRPKM$RPKM)
plot(kernel_d, lwd = 2, main = "Kernel density plot without Log-trans")
```

###The distribution is very skew to left side of the plot.

*Plot the kernel density estimate for log(RPKM+1) (x axis). Describe the distribution.*

```{r}
kernel_d_log<-density(log(rnaSeqDataMeltedRPKM$RPKM+1))
plot(kernel_d_log, lwd = 2, main = "Kernel density plot with Log-trans")
```

###This time it is better to see the distribution, there are two peaks at around 0.1 and 2.5.

*Why do we log-transform gene expression data?*
###Because the original RPKM data have very large variance, some values are very large, this would cause the small values to be hard to see, thus we can not see the distribution clearly. Log-transform would reduce this variance while keep the difference between each values, and we can see the distribution clearly.

*Why do we use RPKM+1?*
###Because some values are 0, this would cause error when doing log transform, and RPKM+1 would finally result in 0, which is equal to the original one. Indeed, most of the RPKM values are larger than 1, values lower than 1 is not as such important.

#### Question 1e ( 2pts)
*It is common to exclude some genes from analysis. For instance, if the gene is not expressed, there is no way it can be differentially expressed. Even in cases where there are a few reads, it is common to exclude these genes too since there we cannot be confident in any differential expression based on very few reads. Calculate the mean number of reads per gene using the count data. We're going to use a cutoff of a mean of at least 5 reads per gene, on average, to be included in the subsequent analysis. Plot the mean reads per gene as a Cumulative Distribution (see https://en.wikipedia.org/wiki/Cumulative_distribution_function), and indicate with a vertical line where the cutoff lies.*


```{r}
average_counts<-rnaSeqDataMeltedCounts %>% group_by(Gene.ID) %>% summarize("Average_counts"=sum(counts)/10)
plot(ecdf(average_counts$Average_counts),xlim=c(0,100))
abline(v=5,col = "red")
```

*By looking at the graph, what fraction of genes will be excluded if we require an average of at least 5 reads per gene?*
###By looking at the graph, nearly 32% of genes will be excluded.

#### Question 1f
*Make a vector of `Gene.ID`s for all the genes we will _include_ in our analysis (e.g. those with 5 or more reads per sample on average). Name it `keepGenes`.*

```{r}
keepGenes<-average_counts$Gene.ID[average_counts$Average_counts>5]
```

*How many genes have we retained?*

```{r}
length(keepGenes)
```
###We retain 15537 genes. 

#### Question 1f ( 2 bonus pts)

*`rnaSeqDataMeltedCounts` contains a commmon error introduced by pasting the data having passed through excel. Can you identify this error and show some examples?*

```{r}
###rnaSeqDataMeltedCounts looks very normal, I cannot find any error.
```


### Question 2

*We want to know what genes are differentially expressed between LCLs and B cells.* 

*Today, we're going to use edgeR. The user guide is here:* https://www.bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf

#### Question 2a (2pts)
*For calculating differential expression, we need a count matrix. There is a lot more information in counts than in normalized counts. For example, you can believe 1/1E6 reads a lot less than 100/1E8 reads, but they both have the same RPKM values. Accordingly, Please create a matrix of counts (genes in the rows and samples in the columns) and filter it for genes with an average number of counts across samples >5 (as in Q1f). *

```{r}
countMatrix = cast(rnaSeqDataMeltedCounts, Gene.ID ~ sampleID, value="counts", fun.aggregate="sum")
row.names(countMatrix) = countMatrix$Gene.ID
countMatrix$Gene.ID=NULL
countMatrix= as.matrix(countMatrix)
countMatrix<-countMatrix[rowSums(countMatrix)/10>=5,]
```

#### Question 2b (4pts)

*EdgeR is exceptionally versatile, with many different options for analysis. Today, you're going to use the GLM-quasi-likelihood approach to calculate differential expression. Below is some incomplete code that you can use to start with. Please complete the code so that edgeR uses your countMatrix to calculate differential expression comparing B cells to LCLs. Please also comment the code so that it is clear what each step does (in your own words).*

```{r}
###get the group information
ct=as.factor(gsub("([A-Z]+)_[ABCDE]","\\1",colnames(countMatrix)))
###produce DGElist
res= DGEList(counts=countMatrix, group=ct)
###since different samples have different coverage, so we need calculate the normalization factor
res = calcNormFactors(res)
###then get the design matrix
design = model.matrix(~ct)
###dispersion estimation
res= estimateDisp(res,design)
###perform quasi-likelihood F-tests
fit= glmQLFit(res,design)
qlf = glmQLFTest(fit, coef=2)
###get all differential status of genes
allDEStats_unpaired = as.data.frame(topTags(qlf,n=nrow(countMatrix)))
allDEStats_unpaired$Gene.ID=row.names(allDEStats_unpaired)
head(allDEStats_unpaired)
```


#### Question 2c (2pts)
*Now, let's confirm that we did the comparison in the correct direction. We want to ensure that we did LCLs/B-cells, and not the other way around. Make a volcano plot using ggplot2 for this analysis. A volcano plot is a scatter plot with log(fold-change) on the x-axis, and -log(P-value) on the y-axis. Label TNF in this plot. Alter the above code if logFC is not expressed as log(LCLs/B-cells). *

```{r}
allDEStats_unpaired$diffexpressed <- "NO"
allDEStats_unpaired$diffexpressed[allDEStats_unpaired$logFC > 3 & allDEStats_unpaired$FDR < 0.01]<-"Up"
allDEStats_unpaired$diffexpressed[allDEStats_unpaired$logFC < (-3) & allDEStats_unpaired$FDR < 0.01]<-"Down"
allDEStats_unpaired$delabel <- NA
allDEStats_unpaired$delabel[rownames(allDEStats_unpaired)=="TNF"]<-"TNF"
ggplot(data=allDEStats_unpaired, aes(x=logFC, y=-log10(PValue), col=diffexpressed, label=delabel))+geom_point()+theme_minimal()+geom_text()+scale_color_manual(values=c("blue", "black", "red"))
```
###Since I have corrected the labels, so this time the TNF should show in up-regulated genes, so logFC is correctly expressed as log(LCLs/B-cells).

#### Question 2d (2pts)
*In the case of this experiment, each donor had corresponding B cell and LCL samples - these are indicated by the 'donor' variable, which takes the values A-E. For instance, the LCLs and B-cells labeled with Donor 'E' are both from the same individual. What happens if there are person-specific differences in expression? For instance, what if person E had autoimmune disease? This could perturb both their B-cells and their resulting LCLs. These are called 'paired samples' since there is a pair of samples for each donor (B cells and LCLs). Repeat the analysis above, this time incorporating the knowledge of paired samples in the analysis, using the donor as a blocking variable.*

```{r}
donor<-as.factor(gsub("([A-Z]+)_([ABCDE])","\\2",colnames(countMatrix)))
design = model.matrix(~donor+ct)
res= estimateDisp(res,design)
fit= glmQLFit(res,design)
qlf = glmQLFTest(fit, coef=6)
allDEStats_paired = as.data.frame(topTags(qlf,n=nrow(countMatrix)))
allDEStats_paired$Gene.ID=row.names(allDEStats_paired)

allDEStats_paired$diffexpressed <- "NO"
allDEStats_paired$diffexpressed[allDEStats_paired$logFC > 3 & allDEStats_paired$FDR < 0.01]<-"Up"
allDEStats_paired$diffexpressed[allDEStats_paired$logFC < (-3) & allDEStats_paired$FDR < 0.01]<-"Down"
allDEStats_paired$delabel <- NA
allDEStats_paired$delabel[rownames(allDEStats_paired)=="TNF"]<-"TNF"
ggplot(data=allDEStats_paired, aes(x=logFC, y=-log10(PValue), col=diffexpressed, label=delabel))+geom_point()+theme_minimal()+geom_text()+scale_color_manual(values=c("blue", "black", "red"))
```

### Question 3
*When performing any bioinformatic analysis, it is critically important to verify your answers. Here, we're going to compare the results from our paired analysis vs. the unpaired analysis to ensure that our analysis is correct (go back and redo the previous parts if you find that your analysis appears to be incorrect).*

#### Question 3a ( 1pts)
*Compare the results from the paired analysis vs. the original analysis (where donor was not considered).*
*How many genes have a logFC that differs by 1 or more?*

```{r}
diff_indices <- c()
orderpaired <- allDEStats_paired[order(row.names(allDEStats_paired)), ]
orderunpaired <- allDEStats_unpaired[order(row.names(allDEStats_unpaired)), ]
lengthlog <- 0
for(i in 1:nrow(orderpaired)){
  
  if(abs(orderunpaired$logFC[i] - orderpaired$logFC[i]) >= 1){
    
    lengthlog = lengthlog +1
    diff_indices <- c(diff_indices,i)
    
  }
} 


print(lengthlog)
```

#### Question 3b ( 3pts)
*Plot the logFC values for paired (x-axis) vs unpaired (y-axis) analyses. What is the Pearson r between the two?*

```{r}

Unpaired <- orderunpaired$logFC
Paired <- orderpaired$logFC
plotted <- plot(Paired, Unpaired)
correlation2<-cor(Paired,Unpaired, method = "pearson")
correlation2

```

#### Question 3c 
*Why are there some genes whose logFC estimates differ so much between the paired and unpaired analyses? Let's inspect the genes with a logFC difference between paired and unpaired of at least 1. A heatmap can be used to display a matrix of values. In this case, we want the values to represent log(RPKM+1) values (use `rnaSeqDataMeltedRPKM`), with genes (only those whose logFCs differ between the analysis using donor as a blocking variable vs. without by more than 1) on the y axis and samples on the x axis.* 
  
```{r}
#rnaSeqDataMeltedRPKM$RPKM

genegreater <- vector(mode = "character", length = 88)
rpkmgreater <- vector(mode = "character", length = 88)
#filteredlog <- vector(mode="numeric", length = length(rnaSeqDataMeltedRPKM$Gene.ID))
#rpkmordered <- rnaSeqDataMeltedRPKM[order(row.names(rnaSeqDataMeltedRPKM)), ]
#rpkmnew = cast(rpkmordered, Gene.ID ~ sampleID, value="RPKM", fun.aggregate="sum")
for(i in 1:88){
  ind<-diff_indices[i]
  genegreater[i] <- allDEStats_paired$Gene.ID[ind]

}

rpkmgreater <- rnaSeqDataMeltedRPKM[which(rnaSeqDataMeltedRPKM$Gene.ID %in% genegreater), ]
rpkmgreater$logRPKM <- log(rpkmgreater$RPKM + 1)

ggplot(rpkmgreater, aes(sampleID, Gene.ID, fill = logRPKM)) + geom_tile() + scale_fill_gradient(low="white", high="blue")

#logmat <- matrix(0, nrow = 23228, ncol = 11)
#colnames(logmat) <- c("Gene.ID", "B_A", "B_B", "B_C", "B_D", "B_E", "LCL_A", "LCL_B", "LCL_C", "LCL_D", "LCL_E")
#rownames(logmat) <- rownames(rpkmnew)
#logmat[,1] <- rpkmnew$Gene.ID 
#for(i in 1:length(rpkmordered))
#  for( n in 2:length(rpkmnew))
#    if (abs(orderunpaired$logFC[i] - orderpaired$logFC[i]) >= 1){
#      logmat[i,n] <- rpkmnew[i,n] 
#    }else{
#      logmat[i,n] <- 0
#    }
#lengthlog2

#lengthlog
#rpkm<-log(rpkmMatrix+1)
#pheatmap(rpkm)

```

*What are some reasons why paired and unpaired analyses appear to be giving different answers in this case? Are there donor-specific things going on?*
  
No donor specific things going on. Since the paired analysis removed any possible repetitions in the analysis, we then presumably get lower values for a majority of the readings.

  
### Question 4
*There are many different ways of normalizing RNA seq data and they all make different assumptions. *
  
*For more information on RNA-seq normalization methods, see this great post by Harold Pimentel: https://haroldpimentel.wordpress.com/2014/05/08/what-the-fpkm-a-review-rna-seq-expression-units/*
  
#### *From now on, we will be using only the paired (donor as blocking variable) analysis*
  
#### Question 4a ( 3.5 pts)
  
  *edgeR uses TMM normalization. TMM normalization is usually a solid approach, but it assumes that most of the genes do not differ in expression between your samples. One approach you could use to check this is by looking at the P-value distribution from our DE analysis. For a well-behaved statistical test, P-values are uniformly distributed between 0 and 1. Plot the Cumulative Distribution of P-values for the paired analysis. On the same plot, include a CDF for 100,000 uniformly sampled random values between 0 and 1.*
  
```{r}
rand.unif <- runif(100000, min = 0, max = 1)
plot(ecdf(orderpaired$PValue), main="P-Value and CDF")
par(new = TRUE)
plot(ecdf(rand.unif), main = " ") 
```

*Do the P-values calculated by edgeR look like they are sampled uniformly between 0 and 1?*
  
  *Estimate the fraction of genes that appear to have differential expression (does not have to be exact):* 
  
  *Do you think we should use TMM normalization for these samples? Why or why not?*
  
#### Question 4b (3.5 pts)
  *Another way to ensure that P-values are well-behaved is to make a Q-Q plot. A Q-Q plot, in this case, will show the expected P-values (under the null hypothesis) on the x axis, and actual p-values on the y-axis. Recall that by the definition of a P-value, if you generate `n` p-values under the null hypothesis, you expect to see 1 p-value of 1/`n` or lower, two of 2/`n` or lower, etc. Here, `n` is the number of DE tests you performed. Make a Q-Q plot of p-values for your DE analysis. Use -log(P) for both axes (this puts the most significant in the upper right, and least significant in the lower left). Also include the line y=x in red for comparison. For a well-behaved statistical test, most of the points should lie along the y=x line. *
  
  
```{r}
ggplot(allDEStats_paired,aes(x = -log(seq(1/length(keepGenes), 1, by = 1/length(keepGenes))), y = -log(PValue))) + geom_abline(a = 0, b = 1, color = 'red') +geom_point(size = 1) + xlab("-log(expected pvalues)") + ylab("-log(paired pvalues)") + ggtitle ("Q-Q plot using expected and actual p-values")

```

*Do most of the P-values lie along the y=x line?*
  
  No, most of the pvalues are above the y=x line.
  
  *What does this mean about how we did our analysis? What would you do differently if we were to redo the analysis?*
  
  Since the expected pvalues and actual pvalues are not in line, we can decide that our pvalues are biased to certain parts of our data. We may need to redo analysis and more accurately consider assumptions made during analysis.
  
  #### Question 4c (3 pts)
  *FDRs are automatically calculated by edgeR. plot the -log10(FDR) (y axis) versus the -log10(P-value) (x axis) for the paired DE analysis. Include the line y=x in red again for reference*
  
```{r}
ggplot(allDEStats_paired,aes(x = -log10(PValue), y = -log10(FDR))) + geom_abline(a = 0, b = 1, color = 'red') + geom_point(size = 1) + xlab("-log(paired pvalues)") + ylab("-log(FDR)") + ggtitle ("Q-Q plotting paired FDR and p-values")
```

*Are the points above or below the line? Why?*
  The points are below the line, but much closer to the actual pvalues because the FDR values corrects the multiple tests. 
  
#### Question 4d (3 pts)
*If we took our set of significant DE genes to be those that meet an FDR of 10%, how many genes would be considered significantly DE?*
  
```{r}

```

*What is the P-value of the gene with the highest FDR that is still less than or equal to FDR < 10%?*
  
```{r}

```

*Ignoring any potential flaws in the analysis up to this point and assuming our P-values and FDRs are well founded, is the probability of this gene NOT being differentially expressed (ie. a false discovery) greater than, less than, or equal to 10%?*
  
  
  
### Question 5
*Now we're going to spend some time actually looking at the results of our analysis, trying to determine what is going on when B-cells are converted to LCLs.*

#### Question 5a (2 pts)

*GOrilla, which we use below, can take a single ranked list as input. Here, you should make two ranked lists of gene IDs, one with the gene IDs sorted by log FC from high to low, and the other from low to high. Use the paired analysis for this question. You will only want gene IDs written to the file (e.g. no row or column names, no quotes) - GOrilla will give error messages if it is not properly formatted.*


```{r}
lowtohigh <- allDEStats_paired[order(allDEStats_paired$logFC), ]
hightolow <- allDEStats_paired[order(allDEStats_paired$logFC,decreasing=TRUE), ]
genelowtohigh <- lowtohigh$Gene.ID
genehightolow <- hightolow$Gene.ID
writeLines(genelowtohigh, con = "lowtohigh.txt")
writeLines(genehightolow, con = "hightolow.txt")
```

#### Question 5b (6 pts)
*Now use GOrilla to measure gene set enrichment analysis for both sets of genes.*
http://cbl-gorilla.cs.technion.ac.il/

*Use Homo sapiens; a single ranked list of genes (those we just created - run it twice, once per file); and all ontologies.*

*When inspecting the results, consider both the significance of the enrichment, the degree of enrichment, and the number of genes involved. For instance, a high-fold enrichment for a very specific term can give more insight than a more significant but lower fold enrichment for a more generic term. Alternatively, enrichments involving fewer genes are more likely to occur by chance, and so may have a high fold enrichment, but nonetheless be spurious.  *

*What processes are relatively high in B cells?*
	multicellular organismal process
	regulation of ion transport
	cell adhesion
	biological adhesion
  
*Which are relatively high in LCLs?*
adult heart development
anatomical structure development
cellular development process
chemotaxis
cell differentiation
cell recognition?
signalling pathways

*What do you think could be happening as B-cells are reprogrammed to LCLs?*
The TNF Gene should be highly induced in LCLs to form B-cells again. However for the creation of LCLs from B-cells TNF might be removed from the B-cells